Last glacial cycle glacier erosion potential in the Alps

### 2021-05-24 articles/erosion/response-referee1.md

<!-- Copyright (c) 2021, Julien Seguinot (juseg.github.io), Ian Delaney
Creative Commons Attribution-ShareAlike 4.0 International License
(CC BY-SA 4.0, http://creativecommons.org/licenses/by-sa/4.0/) -->

# Authors' response to Anonymous Referee #1

Dear Anonymous Referee #1,

Thank you very much for your positive comments and for your time spent
reviewing our manuscript.

@@

Thank you for this accurate summary of our paper.

## General comments

> Authors decided to keep the paper short, referring most information on
> glacial model set up and resulting ice cover to the Seguinot et 2018 paper.
> This makes it no always very easy to read and needs checking in the original
> manuscript (for example, ice cover deviations model – field bases LGM
> extension).

Indeed, it is our deliberate choice to avoid paraphrasing or duplicating
content from the 2018 paper, where the preparation of the text and figures also
involved a different author team. We believe this is justified given that both
papers will be published open-access. Nonetheless, small additions were made.
Please see our responses to your technical comments below.

@@

> The introduction gives a nice, relatively long (in relation to other
> chapters) overview on different aspects of glacial erosion and resulting
> features, and guides through problems to investigate them. Here, I don’t see
> very clear how different aspects or parts in the introduction are
> specifically addressed in this study, i.e. which problems are exactly aimed
> to be solved of this large portfolio of mentioned gaps in the understanding
> of glacial processes. I think this could be better formulated and balanced.

TODO: This remain a tricky points to me. Shoulw we remove parts of the
introduction? Or add sentences tying our statements to upcoming results?

@@

> I think the study could also well contribute to the discussion on the
> elevation distribution of the cumulative glacial erosion over several cycles
> in the Alps (high elevations vs low elevation, e.g. Valla et al., 2011). A
> plot showing the integrated glacial erosion potential over (e.g. 100 m)
> elevation bins would easy to do and eventually an enlightening supplement
> summing up Fig. 5a.

Thank you for the suggestion. Figure 5 was amended with (as Fig. 5b) the
distribution of cumulative erosion potential over 100 m elevation bins.
The resulting plot shows that most modelled erosion occurs below 2000 m,
with the potential erosion volumes between 1000 and 2000 m elevation.

@@

> I would also recommend to make some regional statements on the distribution
> of the glacial erosion potential (Fig. 2). It is quite obvious that some
> tectonic massifs can readily be discriminated, e.g. the Tauern Window, Oetzal
> Crystallin complex, the Aare Massif, Mt. Rosa - Gran Paradiso and Pelvoux
> Massif. Is it the steepness of these low erodible crystallin massifs making
> the erosional potential appearing strikingly high? I think this should/could
> be discussed..

Our calculations indeed produce locally high cumulative erosion potential in
the aforementioned crystalline massifs. More accurate modelling of small-scale
glaciers would be needed to answer your question. Steep topography, though, is
likely to be part of the explanation. A more substantial discussion of the high
erosion potential values in these areas and the model limitations that hinder
interpretation was added in the "age of the glacial landscape" subsection.
The new passage reads as follow:

    The validity of the model results at high elevation is discussable.
    Cristalline massifs such as the Ecrins, Gran Paradiso, Monte Rosa, Aare,
    \"Otztal and Tauern Massifs locally exhibit a strikingly high erosion
    potential. However, the computation of glacier flow velocities on such
    steep surfaces is strongly limited by the model horizontal resolution of
    1\,km, the shallow-ice glacier flow physics (Imhof et al., 2019), and
    PISM's current mass-conservation heuristics (Imhof, 2021). Besides,
    bergschrund (rimaye) processes likely to dominate interglacial cirque
    erosion at such altitudes (Sanders et al., 2012) are not captured by the
    velocity-based glacier erosion power-laws.

@@

> Line 24 Maybe a bit odd to use 4 very old references from the Alps only (you
> hardly used more than 2 references throughout the MS and there are also high
> variety of glacial landforms in other mt ranges as you mentioned).

We agree that the references are not directly relevant. The citation was
reduced to the latest reference by Penck (1905) and moved within the sentence
to clarify that it is specific to the Alps.

> Line 33 I think that’s not very easy to understand what landform you refer
> to? The “periglacial blockfields topped by glacial erratic boulders” (Wirsig
> et al., 2018)? Eventually be more precise on this would help readers.

Agreed. "Characteristic landform preservation" was replaced with "preserved
periglacial blockfields topped by erratic boulders".

@@

> Line 101 Could be misleading as (glacial and periglacial) cirque erosion
> processes are not really covered by any glacial erosion law discussed here
> (or any other I am aware of, Sanders et al., 2012).

Thank you for pointing out this study. This particular sentenced was reworked
to acknowledged not only the limiting resolution but also the lack of relevant
erosion erosion processes. The reference was additionally incorporated in the
relevant part of the discussion (see general comments above).

@@

> Line 127 Higher precipitation increases ice flux and thus erosion, I guess?
> Would eventually helpful to mention (even though it might be referred in
> Seguinot et al., 2018).

Indeed. The increase in ice discharge was made explicit.

> Line 140 Can you be more precise what you mean by realistic? You mean because
> of the localized pattern? Maybe I am wrong but shouldn’t the erosion
> potential at least at the lake Constance overdeepening (Fig. 7) not in the
> order of hundreds of meters (cf. Preusser et al., 2010) and the best fit
> rather (b) or (c) – at least from what I read from the transect (e-h)?
> Visually (a)-(d) seem not to correspond to (e)-(h) if (a)-(d) is also
> presented in meters (annotation at the bar is missing). For example, in (b)
> the maximum erosion potential is like >>1000 m (if in meters) while in
> corresponding (f) it seems clearly lower than 1000 m.

TODO: Another tricky point. Here we need to discuss what is realistic. Probably
no single law gives realistic results indeed.

@@

There is no mismatch between panels (a-d) and (e-h), but to improve readability
we have reduced the number of colour levels on the maps, and changed the
positions of ticks on the transects.

> Furthermore, isn’t the impression of the fit potentially very dependent on
> the initial model parametrization (Seguinot et al., 2018), i.e. the ice flux
> velocity? You should address these dependences!

A paragraph was added under "choice of erosion law" to discuss the
uncertainties on glacier sliding physics. Also including suggestions from
reviewer #2, the new paragraph reads as follow:

    The modelled pattern of erosion potential depends on PISM glacier physics
    and sliding model parameters. In particular, the pseudo-plastic sliding law
    exponent (q=0.25 in Seguinot et al., 2018) controls the sharpness of the
    transition between adherent and decoupled basal conditions. Recent ensemble
    validation of Antarctic Ice Sheet glacial-cycle simulations against
    geological and present-day observations (Albrecht et al., 2020a, b) support
    a higher value of q=0.75, and thus a sliding law closer to linear, which
    would perhaps result in a smoother distribution of sliding velocity and
    erosion potential. Lateral stress gradients missing from the shallow-shelf
    approximation stress balance could also contribute to moderate sliding
    velocity in narrow troughs (Herman et al., 2011; Egholm et al, 2012a, b;
    Pedersen et al., 2014).

@@

> Line 163 Observing Alpine topography I find this result important, which can
> maybe also serve as explanation why e.g. (low erodible) areas away from the
> big troughs covered during glacial maxima only, do surprisingly often show
> no/very low degree in glacial modification (e.g. Ticino; Kelly et al., 2004).

Thank you for pointing this out. The following sentence was added in this
paragraph:

    Nevertheless, it may explain why some areas covered during glacial maxima
    only, such as some valleys on the southern side of the Alps, appear to have
    experienced only little glacial modification of their topography.

@@

> Line 187 “time-transgressive radial pattern”. I don’t understand what you
> mean..

This sentence was only meant to recall the aforementioned results. We rephrased
to "the potential erosion patterns experience spatial shifts through the
glacial cycle".

@@

## Figures

> Fig. 2 Please indicate the outline of the connected ice stream network during
> the LGM (like in Fig. 4 of Seguinot et al., 2018). Especially in the eastern
> and SW Alps coverage largely deviates from what has been suggested from field
> data compilation (e.g. Ehlers and Gibbard, 2004). Even though outcomes might
> not be changing much, it might be helpful to know how much the %overlap is -
> erosional potential might change as fluvial topography turns into glacial one
> (e.g. Harbor, 1988) and this is probably not what you want to mix, I guess.

The LGM outline was added on Fig. 2.

TODO: I still need to add a sentence or two discussing the discrepancy under
"climate sensitivity". I don't understand the point about fluvial topography.

### 2021-05-24 articles/erosion/alps-erosion.tex
Polish response #1, some points still missing.

    In the absence of other landforms, glacially eroded topography has
    sometimes been used as a proxy for glacier extent, for instance in
    geomorphological mapping \citep[e.g.,][]{Margold.etal.2011, Fu.etal.2012}.
    Cold-based glaciers, however, have been observed to preserve landforms as
    fragile as sand beaches \citep{Kleman.1994}, leading to the
    reinterpretation of complex glacial landscapes within a binary conceptual
    framework of cold and temperate basal thermal regimes
    \citep{Kleman.etal.2008, Kleman.etal.2010, Fabel.etal.2012, Fu.etal.2013}.
    Nevertheless, it remains unclear how sharp the erosional
    transition to cold-based ice is, and why some glaciated regions show
    preserved periglacial blockfields and others, including the Alps, do not
    \citep{Wirsig.etal.2016, Seguinot.etal.2018}.

@@

    The modelled cumulative (time-integrated) glacial erosion potential
    (Fig.~\ref{fig:cumulative}a) varies by several orders of magnitude
    from insignificant to hundred-metres-scale erosion potential. Its spatial
    patterns show a very strong localization along the Alpine valleys, with
    local maxima occurring both at
    the Alpine gates where ice flow transited from valley to piedmont glaciers,
    and further upstream where the valley slopes increase. While mountain
    cirques erosion processes may not be captured by the model's physics
    \citep[cf.][]{Sanders.etal.2012} and horizontal resolution, high
    cumulative erosion potential also occurs near the valley heads.
    There is a general tendency for higher cumulative erosion in the
    north-western Alps where the input winter precipitation is higher
    \citep[WorldClim, Fig.~1h in][]{Seguinot.etal.2018} and the glacial relief
    more pronounced in the topography.

@@

    The validity of the model results at high elevation is discussable.
    Cristalline massifs such as the Ecrins, Gran Paradiso, Monte Rosa, Aare,
    \"Otztal and Tauern Massifs locally exhibit a strikingly high erosion
    potential. However, the computation of glacier flow velocities on such
    steep surfaces is strongly limited by the model horizontal resolution of
    1\,km, the shallow-ice glacier flow physics \citep{Imhof.etal.2019}, and
    PISM's current mass-conservation heuristics \citep{Imhof.2021}. Besides,
    bergschrund (rimaye) processes likely to dominate interglacial cirque
    erosion at such altitudes \citep{Sanders.etal.2012} are not captured by the
    velocity-based glacier erosion power-laws. Nevertheless,
    the lack of rapid erosion at high elevations for much
    of the modelled glacial cycle implies that the highest mountain
    cirques such as Ch\"uebodensee (Fig.~\ref{fig:landscape}) would have formed
    over interglacial periods, when glaciers are confined to high terrain
    \citep{Barr.etal.2017, Barr.etal.2019}. Similar processes also occur on
    Himalayan glaciers and may limit erosion for the high elevation portion of
    that mountain range during cold periods as well, failing to offset uplift
    rates \citep{Harper.Humphrey.2003}.

>>>>

    In the absence of other landforms, glacially eroded topography has
    sometimes been used as a proxy for glacier extent, for instance in
    geomorphological mapping \citep[e.g.,][]{Margold.etal.2011, Fu.etal.2012}.
    Cold-based glaciers, however, have been observed to preserve landforms as
    fragile as sand beaches \citep{Kleman.1994}, leading to the
    reinterpretation of complex glacial landscapes within a binary conceptual
    framework of cold and temperate basal thermal regimes
    \citep{Kleman.etal.2008, Kleman.etal.2010, Fabel.etal.2012, Fu.etal.2013}.
    Nevertheless, it remains unclear how sharp the erosional
    transition to cold-based ice is, and why some glaciated regions show
    preserved periglacial blockfields topped by erratic boulders and others,
    including the Alps, do not \citep{Wirsig.etal.2016, Seguinot.etal.2018}.

@@

    The modelled cumulative (time-integrated) glacial erosion potential
    (Fig.~\ref{fig:cumulative}a) varies by several orders of magnitude
    from insignificant to hundred-metres-scale erosion potential. Its spatial
    patterns show a very strong localization along the Alpine valleys, with
    local maxima occurring both at
    the Alpine gates where ice flow transited from valley to piedmont glaciers,
    and further upstream where the valley slopes increase. While mountain
    cirque glaciers and relevant erosion processes may not be captured by the
    model physics and horizontal resolution, high
    cumulative erosion potential also occurs near the valley heads.
    There is a general tendency for higher cumulative erosion in the
    north-western Alps where the input winter precipitation is higher
    \citep[WorldClim, Fig.~1h in][]{Seguinot.etal.2018} and the glacial relief
    more pronounced in the topography.

@@

    The validity of the model results at high elevation is discussable.
    Cristalline massifs such as the Ecrins, Gran Paradiso, Monte Rosa, Aare,
    \"Otztal and Tauern Massifs locally exhibit a strikingly high erosion
    potential. However, the computation of glacier flow velocities on such
    steep surfaces is strongly limited by the model horizontal resolution of
    1\,km, the shallow-ice glacier flow physics \citep{Imhof.etal.2019}, and
    PISM's current mass-conservation heuristics \citep{Imhof.2021}. Besides,
    bergschrund (rimaye) processes likely to dominate interglacial cirque
    erosion at such altitudes \citep{Sanders.etal.2012} are not captured by the
    velocity-based glacier erosion power-laws. Nevertheless,
    the lack of rapid erosion at high elevations for much of the modelled
    glacial cycle (Fig.~\ref{fig:hypsogram}) implies that the highest mountain
    cirques such as Ch\"uebodensee (Fig.~\ref{fig:landscape}) would have formed
    over interglacial periods, when glaciers are confined to high terrain
    \citep{Barr.etal.2017, Barr.etal.2019}. Similar processes also occur on
    Himalayan glaciers and may limit erosion for the high elevation portion of
    that mountain range during cold periods as well, failing to offset uplift
    rates \citep{Harper.Humphrey.2003}.
